Incorporating periodic variability in hidden Markov models for animal movement

### 2016-02-29 paper3.Rnw
back to you ben!

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in latent/hidden behavioural state/mode transition 
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in behavioural transitions over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

For movement behaviours that change on a fast time scale, such
that movement behaviours recorded
at successive observations are effectively independent,
\emph{finite mixture models} (FMM) --- which can be considered
a special case of HMMs where the
probability of state occupancy is independent of the previous state ---
can adequately describe movement \citep{tracey_mapping_2012}.  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities change deterministically over time.
Thus FMMs and HMMs,
with or without temporal variation in the occupancy or transition
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.
\todo{do we want a figure showing this?}

@@

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission staff using
trained hounds and houndsmen. Of these animals, 13 had 
sufficient data to be used by \cite{kerk2015hidden}.
\todo{not sure quite where we should use {\tt citeauthor} vs. {\tt cite} \ldots}
Here we focus
on the four cats with the most data (all with approximately 10,000-15,000 observations: see Table~1 in Supplementary Material), 
in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is toward larger data sets.  As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\citep{turchin1998quantitative,kerk2015hidden}.

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which makes precisely this assumption.
\todo{could say here (or maybe better in the conclusions) that
revisiting HSMM in this context would be worthwhile future work. 
ML: conclusion is good, I still need to add viterbi paragraph}

\cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (step length plus turning angle)
observations.  However, we do briefly compare log-Normal and Weibull
step-length distributions\todo{have we run Gamma as well?}, with and without 
a von Mises-distributed turning angle included in the model (Figure~\ref{fig:BICred_plot}).\todo{W/VM/HMM (blue/solid/left panel) should correspond to vdK's results, but that has $\dbic=0$ for 5 states rather than 6. Is that right? Is that HMM/HSMM diff?}
(Note that most movement analyses, including
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as independent observations while neglecting possible correlations
between them.)

@@

For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.  The FMM is a special case of HMM where the probabilities of \emph{entering} a given state are identical across all states --- i.e., the probability of occupying a state at the next time step is independent of the current state occupancy. It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$. \todo{BMB: is this right?}

@@

Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity.

\todo{Discuss model complexity/parameter-number computations here, or with each model.  Also discuss the ``hourly'' model here.  Make the table into a figure!}

@@

\todo{Give more details about parameters, structure of simulations, etc.}

@@

\todo{describe the Viterbi algorithm here, and the autocorrelation function
computation - everything that appears in the results should be described
explicitly here....ML: at the bottom}

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($>  $ observations). (1) BIC relative to the optimal-BIC model within each type of transition complexity\todo{BMB: not sure what that means. Clarify?} ($\dbic$ = BIC - min(BIC)). 
(2) Computed average step-length by hour of day for the observed data and for data simulated from the models.
(3) Computed temporal autocorrelation function (ACF) for the observed data and for data simulated from the models. For the second and third metrics (hourly average and ACF), we simulated data from the models because the metrics are hard to compute from first principles for HMMs with temporally varying transition probabilities.

\todo{give at least a hint of why/what each of these three 
approaches ($\dbic$, hourly averages, ACF) will show. ML: What do you think?}

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#130.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
load("cat1allmod.RData")
df2 <- transform(allmodsumdf,
            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
            turn_angle=ifelse(grepl("VM",model),"yes","no"),
            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

@@

As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions to compare the temporal effects in goodness of fit. Looking at Figure 2, models with temporal heterogenity is better than models without in both FMM and HMM frameworks, but time homogenous HMM models is better than FMM with sinusoidal temporal heterogeneity. Temporal heterogenity is good with good models such as the quadratic and sinuosidal. Multiple block models is good for cats with distinguishable diurnal patterns and hourly model is too complicated.

@@

<<state_plots cat2, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
load("cat2sumdat.RData")
temp5 <- tail(getpars(fithomo5s),10)
temp4 <- tail(getpars(fitsin4s),8)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
dat2 <- data.frame(log_mean=temp4[c(1,3,5,7)],
                     log_sd=temp4[c(2,4,6,8)],Model="HMM + TH(sin)") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

@@

<<state_plots cat14, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
load("cat14sumdat.RData")
temp5 <- tail(getpars(fithomo5s),10)
temp4 <- tail(getpars(fitsin4s),8)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
dat2 <- data.frame(log_mean=temp4[c(1,3,5,7)],
                     log_sd=temp4[c(2,4,6,8)],Model="HMM + TH(sin)") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

@@

<<state_plots cat15, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="States">>=
load("cat15sumdat.RData")
temp5 <- tail(getpars(fitsin5s),10)
temp6 <- tail(getpars(fithomo6s),12)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM + TH(sin)") 
dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
                     log_sd=temp6[c(2,4,6,8,10,12)],Model="HMM") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

@@

The biggest issue with multiple states is identifying them biologically. We really do not know if there Florida panthers really moves according to hidden states, and even if they do, we don't know the true number nor able to observe them. Three distinct movement states seems biologically interpretable for Florida panthers: Short step length suggests resting states, intermediate step length might suggest foraging state and long step length might suggest traveling state. Long step lengths are easy to capture in all models, where as short step lengths are hard to capture given the GPS errors and unstable movements patterns suggesting multiple states associate with short step lengths. Thus, lower number of short step length states with higher variance can be equivalent or more realistic than multiple distinguishable short step length states with low variance.

@@

\appendix
\section{Supplementary material}
\begin{table}[]
\centering
\caption{Cat-ID}
\label{Cat-ID}
\begin{tabular}{ccc}
  (van de Kerk 2015) & (IR@UF)  & Number of Observations  \\
  130 & 1   & 10286    \\
  131 & 2   & 9458    \\
  48  & 14  & 14645    \\
  94  & 15  & 10250  
\end{tabular}
\end{table}

@@

<<BICred_plot cat131,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#131.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
allmodsumdf <- readRDS("cat2allmod.RDS")
df2 <- transform(allmodsumdf,
            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
            turn_angle=ifelse(grepl("VM",model),"yes","no"),
            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

<<BICred_plot cat94,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#94.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
allmodsumdf <- readRDS("cat15allmod.RDS")
df2 <- transform(allmodsumdf,
            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
            turn_angle=ifelse(grepl("VM",model),"yes","no"),
            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

<<viterbi, warning=FALSE, message=FALSE, echo=FALSE >>=
load("cat1.RData")
vitcat <- function(catt,mod){
  params <- tail(getpars(mod),6)
  vitdist <- numeric(ntimes(mod))
  states <- viterbi(mod)$state
  for(i in 1:ntimes(mod)){
    if(states[i] == 1){
      vitdist[i] <- rnorm(1,params[1],params[2])
    }
    if(states[i] == 2){
      vitdist[i] <- rnorm(1,params[3],params[4])
    }
    if(states[i] == 3){
      vitdist[i] <- rnorm(1,params[5],params[6])
    }
  }
  dat <- data.frame(viterbi=vitdist,time=catt$Time, obs=catt$LogDist)
  return(dat)
}

>>>>

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in latent/hidden behavioural state/mode transition 
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in behavioural transitions over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales \citep{gurarie2009novel}.

For movement behaviours that change on a fast time scale, such
that movement behaviours recorded
at successive observations are effectively independent,
\emph{finite mixture models} (FMM) --- which can be considered
a special case of HMMs where the
probability of state occupancy is independent of the previous state ---
can adequately describe movement \citep{tracey_mapping_2012}.  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities change deterministically over time.
Thus FMMs and HMMs,
with or without temporal variation in the occupancy or transition
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

@@

GPS collars were fitted to 18 Florida panthers in 2005-2012 by Florida
Fish and Wildlife and Conservation Commission staff using
trained hounds and houndsmen. Of these animals, 13 had 
sufficient data to be used by \cite{kerk2015hidden}.
Here we focus
on the four cats with the most data (all with approximately 10,000-15,000 observations: see Table~1 in Supplementary Material), 
in part because our goal is to understand
the issues that arise when simple models are fitted to large data sets, and 
in part because the general trend in telemetry studies is toward larger data sets.  As is typical in studies of animal movement, 
we took first differences of the data by decomposing
contiguous sequences of hourly GPS coordinates into 
successive step lengths (in meters) and turning angles (in radians)
\citep{turchin1998quantitative,kerk2015hidden}.

\cite{kerk2015hidden} used hidden semi-Markov models (HSMM), 
an extension of HMM that permits explicit modelling of dwell times 
\citep{langrock_flexible_2012}, considering both Poisson and
negative binomial distributions for dwell times.  As shown 
by \cite{kerk2015hidden} (Figure S3b, top row, middle panel),
the estimated shape parameter of the negative binomial dwell time
distribution was typically close to 1 ($\approx 0.4-1.6$; confidence
intervals were not calculated), implying that a geometric distribution
(i.e., negative binomial with shape=1) might be adequate. In turn,
this suggests that we might not lose much accuracy by reverting to 
a simpler HMM framework, which makes precisely this assumption.

\cite{kerk2015hidden} considered time-homogeneous models with 
a variety of candidate distributions %
--- log-Normal, Gamma, and Weibull distributions for step lengths
and von Mises and wrapped Cauchy distributions for the turning angle --- %
concluding on the basis of the Akaike information criterion (AIC) 
that Weibull step length and wrapped Cauchy turning angle distributions
were best.  
Since our analysis aims for simplicity and qualitative
conclusions rather than for picking the very best
predictive model, we focus on models that treat each step as
a univariate, log-Normally distributed observation, glossing
over both the differences in shape between
the three candidate step-length distributions and the effects
of considering multivariate (step length plus turning angle)
observations.  However, we do briefly compare log-Normal and Weibull
step-length distributions, with and without 
a von Mises-distributed turning angle included in the model (Figure~\ref{fig:BICred_plot}).
(Note that most movement analyses, including
\cite{kerk2015hidden}, are only partially multivariate, 
treating step length and turning angle at a particular time
as independent observations while neglecting possible correlations
between them.)

@@

For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.  The FMM is a special case of HMM where the probabilities of \emph{entering} a given state are identical across all states --- i.e., the probability of occupying a state at the next time step is independent of the current state occupancy. It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$.

@@

Model complexity and number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity. See supplementary material Table 1.

@@

@@

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($>  $ observations). (1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)). 
(2) Compare average step-length by hour of day for the observed data and for data simulated from the models.
(3) Compare temporal autocorrelations for the observed data and for data simulated from the models. For the second and third metrics (hourly average and ACF), we simulated data from the models because the metrics are hard to compute from first principles for HMMs with temporally varying transition probabilities.

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#1.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
load("cat1allmod.RData")
df2 <- transform(allmodsumdf,
            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
            turn_angle=ifelse(grepl("VM",model),"yes","no"),
            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

@@

As a complement, we also added FMM, FMM with sinusoidal temporal heterogeneities, and HMM with hourly transitions (24-hour multiple block transition model) to compare the temporal effects in goodness of fit. Looking at Figure 2, models with temporal heterogenity is better than models without in both FMM and HMM frameworks, but time homogenous HMM models is better than FMM with sinusoidal temporal heterogeneity. Temporal heterogenity is good with good models such as the quadratic and sinuosidal. Multiple block models is good for cats with distinguishable diurnal patterns and hourly model is too complicated.

@@

<<state_plots cat2, warning=FALSE, message=FALSE, echo=FALSE>>=
load("cat2sumdat.RData")
temp5 <- tail(getpars(fithomo5s),10)
temp4 <- tail(getpars(fitsin4s),8)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
dat2 <- data.frame(log_mean=temp4[c(1,3,5,7)],
                     log_sd=temp4[c(2,4,6,8)],Model="HMM + TH(sin)") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

@@

<<state_plots cat14, warning=FALSE, message=FALSE, echo=FALSE>>=
load("cat14sumdat.RData")
temp5 <- tail(getpars(fithomo5s),10)
temp4 <- tail(getpars(fitsin4s),8)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM") 
dat2 <- data.frame(log_mean=temp4[c(1,3,5,7)],
                     log_sd=temp4[c(2,4,6,8)],Model="HMM + TH(sin)") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

@@

<<state_plots cat15, warning=FALSE, message=FALSE, echo=FALSE, fig.cap="State Parameter Estimates">>=
load("cat15sumdat.RData")
temp5 <- tail(getpars(fitsin5s),10)
temp6 <- tail(getpars(fithomo6s),12)
dat1 <- data.frame(log_mean=temp5[c(1,3,5,7,9)],
                     log_sd=temp5[c(2,4,6,8,10)], Model="HMM + TH(sin)") 
dat2 <- data.frame(log_mean=temp6[c(1,3,5,7,9,11)],
                     log_sd=temp6[c(2,4,6,8,10,12)],Model="HMM") 
dat1 <- dat1[order(dat1$log_mean),]
dat2 <- dat2[order(dat2$log_mean),]

@@

The biggest issue with multiple states is identifying them biologically. We really do not know if there Florida panthers really moves according to hidden states, and even if they do, we don't know the true number nor able to observe them. Three distinct movement states seems biologically interpretable for Florida panthers according \citet{kerk2015hidden}: Short step length suggests resting states, intermediate step length might suggest foraging state and long step length might suggest traveling state. Long step lengths are easy to capture in all models, where as short step lengths are hard to capture given the GPS errors and unstable movements patterns suggesting multiple states associate with short step lengths. Thus, lower number of short step length states with higher variance can be equivalent or more realistic than multiple distinguishable short step length states with low variance.

@@

\newpage
\section{Supplementary material}

\begin{table}[]
\centering
\caption{Cat-ID}
\label{Cat-ID}
\begin{tabular}{ccc}
  (van de Kerk 2015) & (IR@UF)  & Number of Observations  \\
  130 & 1   & 10286    \\
  131 & 2   & 9458    \\
  48  & 14  & 14645    \\
  94  & 15  & 10250  
\end{tabular}
\end{table}

@@

<<BICred_plot cat131,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#2.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
allmodsumdf <- readRDS("cat2allmod.RDS")
df2 <- transform(allmodsumdf,
            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
            turn_angle=ifelse(grepl("VM",model),"yes","no"),
            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

<<BICred_plot cat94,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Change in BIC across models, cat \\#15.  $\\dbic$ is scaled individually per model, so that $\\dbic=0$ represents the BIC-optimal number of states for a particular model.",fig.width=6,fig.height=4>>=
allmodsumdf <- readRDS("cat15allmod.RDS")
df2 <- transform(allmodsumdf,
            distrib=ifelse(grepl("LN",model),"log-Normal","Weibull"),
            turn_angle=ifelse(grepl("VM",model),"yes","no"),
            homog=ifelse(grepl("HMM",model),"homogeneous HMM","time-heterogeneous HMM"))
ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))+
    geom_line()+
    facet_wrap(~ homog)+
    scale_colour_brewer(name="step length\ndistribution",palette="Set1")+
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    scale_linetype(name="von Mises\nturning angle?")+
    labs(y=expression(Delta*"BIC"),
         x="Number of States")+
    geom_hline(yintercept=0,colour="gray")+
    zmargin
@

<<viterbi, warning=FALSE, message=FALSE, echo=FALSE >>=
load("cat1.RData")
load("cat1sumdat.RData")
vitcat <- function(catt,mod){
  params <- tail(getpars(mod),6)
  vitdist <- numeric(ntimes(mod))
  states <- viterbi(mod)$state
  for(i in 1:ntimes(mod)){
    if(states[i] == 1){
      vitdist[i] <- rnorm(1,params[1],params[2])
    }
    if(states[i] == 2){
      vitdist[i] <- rnorm(1,params[3],params[4])
    }
    if(states[i] == 3){
      vitdist[i] <- rnorm(1,params[5],params[6])
    }
  }
  dat <- data.frame(viterbi=vitdist,time=catt$Time, obs=catt$LogDist)
  return(dat)
}
